---
title: "PROSPECT logistic regression for erection reliability - 1 year"
author: "Dr. Rianne Fijten"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Set/your/path/here/")

scriptsLocations = paste(getwd(), "scripts/")
source("path/to/scripts/importAllHandmadeFunctions.R")
importAllHandmadeFunctions(scriptsLocations)
library(ggplot2)
allData1year = readRDS("path/to/data/allData1year")
```

# build 1-year prediction models for erectile dysfunction
We will create a logit prediction model for the outcome *reliability*, which is defined as the reliability of the erection, i.e. how often can you get an erection when you'd like an erection.

```{r, create datasets}
trainingSet = allData1year[[1]]
trainingSet = trainingSet[,1:38]

testSet = allData1year[[2]]
testSet = testSet[,1:38]

``` 

Before we start, we'll first remove the variables related to hot flashes and sensitive breasts as these are not to be included upon request by collaborators.

```{r}
trainingSet = trainingSet[,-grep("epic26_22_opvliegers1", names(trainingSet))]
trainingSet = trainingSet[,-grep("epic26_23_gevoeligeborsten1", names(trainingSet))]

testSet = testSet[,-grep("epic26_22_opvliegers1", names(testSet))]
testSet = testSet[,-grep("epic26_23_gevoeligeborsten1", names(testSet))]

```

First we recreate the training and test set with only the input variables and the outcome we want, in this case *"epic26_19_kwanterectie2". We'll also remove the outcome data that we're not interested in right now from the training and test set.

```{r, select outcomes}
outcomeTraining = allData1year[[1]]$epic26_19_kwanterectie2
outcomeTest = allData1year[[2]]$epic26_19_kwanterectie2
```

Then we'll visualize the distributions of this outcome for the training set to decide how to transform it into a binary classification. For this outcome, 1 means that patients could never get an erection, whereas 5 means they could always get an erection when they wanted one.

```{r, pie charts}
pTraining = createPieChart(outcomeTraining, "Erection reliability in training set")
pTest = createPieChart(outcomeTest, "Erection reliability in test set")

multiplot(pTraining, pTest, cols = 2)
```

We'll first remove patients with missing values (answer = 0). The test set also has zeros in the dataset, so we'll remove those too.

```{r}
trainingSet = trainingSet[-which(outcomeTraining == 0),]
outcomeTraining = outcomeTraining[-which(outcomeTraining == 0)]
testSet = testSet[-which(outcomeTest == 0),]
outcomeTest = outcomeTest[-which(outcomeTest == 0)]

pTraining = createPieChart(outcomeTraining, "Erection reliability in training set")
pTest = createPieChart(outcomeTest, "Erection reliability in test set")
multiplot(pTraining, pTest, cols = 2)
```

The dataset shows that 44-47% of patients are never able to get an erection when they want one. Therefore we will try two approaches. First, we'll compare patients without any issues to patients with some issues (ranging in severity, i.e. who answered 1-4). Second, we'll compare patients that can never have an erection (answer = 1) to the others (answer = 2-5).

Overall (for both training and test set) the distribution of the outcome is as follows:

```{r pie chart of all data}
createPieChart(c(outcomeTraining, outcomeTest))
```

We'll now have a look at whether treatments can play a role in the outcome we're investigating

```{r, distribution treatment + urge answers}
adjacencyList = createAdjacencyList(cbind(trainingSet$treatments, outcomeTraining))

nodeNames = c("Prostatectomy", "EBRT", "Brachy", "Active surveillance", "I NEVER had an erection when I wanted one", "I had an erection LESS THAN HALF the time I wanted one", "I had an erection ABOUT HALF the time I wanted one", "I had an erection MORE THAN HALF the time I wanted one", "I had an erection WHENEVER I wanted one")
createSankeyDiagram(adjacencyList, nodeNames)
```

We can see here that most patients that could never get an erection received prostatectomy. Active surveillance had the least amount of people expereincing problems with their erectile functioning. 

## patients that never have an erection vs. all others

Here we will compare patients that can never have an erection and people that still can (sometimes). Patients that answered 1 could never have an erection, whereas patients still had the ability to produce an erection, albeit not necessarily always. So we'll make the ones that answered 1 class 1. and the ones that answered 2-5 class 2.

```{r, binary outcome 2}
binaryOutcomeTraining = createBinaryDataset(outcomeTraining, 1, seq(from = 2, to = 5))
binaryOutcomeTest = createBinaryDataset(outcomeTest, 1, seq(from = 2, to = 5))
```

## Recursive Feature Elimination
Before going on to other methods, if necessary, we'll first explore logistic regression with recursive feature elimination. We'll save this RFE logit model to a file since re-running the RFE will yield (slightly) different results each time. *(The creation of the RFE logit model and its saving have  been commented out for the purpose of creating the Knitted RMarkdown file and a load is added instead.)*

```{r, rfe logistic regression 2}
# glmProfile_NeverErection = performLogitRfe(trainingSet, binaryOutcomeTraining)
# saveRDS(glmProfile_NeverErection, paste(getwd(), "/models/rfeResults_Erections_reliability_1year_neverErection_noHotFlashesSensBreasts.rds", sep = ""))
glmProfile_NeverErection = readRDS(paste(getwd(), "/models/rfeResults_Erections_reliability_1year_neverErection_noHotFlashesSensBreasts.rds", sep = ""))
plot(glmProfile_NeverErection, type=c("g", "o"))
```

Now we'll use the variable importance to determine which model works best. 

```{r}
varImportance = calculateVariableImportance(glmProfile_NeverErection)
allLogitModels = reduceVariablesLogit(trainingSet, binaryOutcomeTraining, varImportance)
{plot(rev(allLogitModels[[2]]$numberOfVariables), rev(allLogitModels[[2]]$AUC), ylim = c(0, 1), pch = 16)
  points(rev(allLogitModels[[2]]$numberOfVariables), rev(allLogitModels[[2]]$sensitivity), col = "red",  pch = 16)
  points(rev(allLogitModels[[2]]$numberOfVariables), rev(allLogitModels[[2]]$specificity), col =  "blue",  pch = 16)
  points(rev(allLogitModels[[2]]$numberOfVariables), rev(allLogitModels[[2]]$overallAccuracy), col = "green",  pch = 16)
  abline(v=12, col = "red")
  legend("bottomright", legend = c("AUC", "sensitivity", "specificity", "overall accuracy"), col = c("black", "red", "blue", "green"), pch = 16)}
```

THe one with 11 variables seems to be the highest up to the maximum we're willing to go. Let's use that one.
The variables in this model are:

```{r}
#chosenModel = 11
#glmProfile_NeverErection_individual = allLogitModels[[1]][[dim(trainingSet)[2] + 1 - chosenModel]]
#saveRDS(glmProfile_NeverErection_individual, "models/rfeResults_Erections_reliability_1year_neverErection_individual_noHotFlashesSensBreasts.rds")
glmProfile_NeverErection_individual = readRDS("models/rfeResults_Erections_reliability_1year_neverErection_individual_noHotFlashesSensBreasts.rds") 
glmProfile_NeverErection_individual$coefficients
```

Here's an overview of the accuracy of the model on the training set.

```{r}
trainingAccuracy = calculateAccuracy(glmProfile_NeverErection_individual, binaryOutcomeTraining)
trainingAccuracy[[1]] # sensitivity/specificity
trainingAccuracy[[2]] # confusionMatrix
createROCCurve(glmProfile_NeverErection_individual, binaryOutcomeTraining)
```

## Prediction onto test set.

```{r}
testAccuracy = calculateAccuracy(glmProfile_NeverErection_individual, binaryOutcomeTest, testSet)
testAccuracy[[1]] # sensitivity/specificity
testAccuracy[[2]] # confusionMatrix
createROCCurve(glmProfile_NeverErection_individual, binaryOutcomeTest, testSet)

```

This model works well. 


# nomogram
As a final step, we create a nomogram based on the model

```{r}
library(rms)
dd = datadist(glmProfile_NeverErection_individual$data)
options(datadist = 'dd')

f = lrm(glmProfile_NeverErection_individual$model$`factor(outcomeToTest)` ~ sCT  + alg_alc + epic26_25_weinigenergie1
  + cardiovascularDisease + epic26_19_kwanterectie1
  + epic26_18_kwalerectie1 + gleason_group
  + hormoneTherapy + diabetes +  treatments, data = glmProfile_NeverErection_individual$data)

#summary(f)
nom = nomogram(f, fun=function(x)1/(1+exp(-x)), # or fun=plogis
                fun.at=c(.001,.01,.05,seq(0,1,by=.1),.95,.99,.999),
                funlabel="Outcome to test Probability", lp=F)
plot(nom, force.label=T, xfrac=0.25, cex.axis=.5, cex.var=0.5, tcl=-0.35, ia.space = 0.1)


```